---
title: "Peer Assesement 1"
#   

---

####This is the document that describes my work for Coursera's Reproductible Research, Assesement 1  
Autor:FSC

####___________________________________________________________________ 

####  

#### 1.- Project creation and data reading and processing


The data file for the assignment has been downloaded to my download folder.
The code segment below will use ProjectTemplate library to create a set if folders for this project, copy the file to the data folder and unzip it, then it reads the data into 'repdata' object
```{r, results = 'hide', message=FALSE}

library('ProjectTemplate')
library(plyr)
library(dplyr)
library(tcltk)
library(ggplot2)
library(lubridate)

```


```{r}
WorkDir <- "/Volumes/LION-DATA/COURSERA /DATA SCIENTIST/REPRODUCIBLE RESEARCH"
setwd(WorkDir)
## check for Peer_Assessement_1 folder and create ir if it doesn´t exist.
if(!file.exists("Peer_Assessment_1"))  create.project("Peer_Assessment_1")
setwd("Peer_Assessment_1")

## Copying a file from a dir to another dir in case ir hasn't been copied before
from <-"/users/Fran/Downloads/repdata-data-activity.zip"
to <- "./data/repdata-data-activity.zip"
if(!file.exists(to))  file.copy(from, to)
                                
setwd("./data")   ## changing to data folder and unzip if neccesary
if(!file.exists("activity.csv")) unzip("repdata-data-activity.zip")
activity <- tbl_df(read.csv("activity.csv"))
```

####_________________________________________________________________________
    
####    

    
####2.- What is mean total number of steps taken per day?
The data frame is cleaned by filtering NA's out, then grouped by date and summarize the total steps into the steps object. 
A histogram  is shown in wich we can see that the mean value in a day is around 11,000 steps.

```{r}
steps <- activity %>%
     filter(steps != "NA") %>%  ## filtering out rows with NA's on steps
     group_by(date) %>% ## grouping by date
     summarise( daySteps = sum(steps)) ## and summing the steps by groupped date.
## histogram 
hist(steps$daySteps,col="red", breaks =seq(0,25000, by = 1000), main = "Total steps per day",, xlab= ("Total daily Steps"), ylab= ("Nr. of occurences"))
```

####  Mean and median  
 The calculated mean and median ( code below) is:  
 
 Mean:      10766.19  
 Median:    10765
 
 So we see that the distribution shows mean symetry ( mean = median). 

```{r}

mean <- summarize(steps, mean(daySteps))
median <- summarize(steps, median(daySteps))
mean
median


```
####_________________________________________________________________________
    
####    

    
####3.- What is the average daily activity pattern?
defining trim_activity as the activity trimmed from NA's, groupped by the interval of the day and summarize the mean of each interval,
then plotting it.

```{r}
trim_activity <- activity %>%
     filter(steps != "NA")  %>% ## filtering out rows with NA's on steps
     group_by(interval) %>%
    summarise( intSteps = mean(steps))
with(trim_activity, {
    par(mar = c(4,4,4,2))
    plot( interval/100, intSteps,type="l",  xlab=" Time of the Day(24 hour)", ylab=" Step activity", ylim = range(intSteps), col ="magenta", xlim = range(0:24), xaxt="n") 
    axis(1, at=c(2,4,6,8,10,12,14,16,18,20,22,24)) 
    title(main = "Average step activity by time inteval")
    
})
```


Then search for the max of the day...

```{r}

maxActivity <- summarise(trim_activity, maxSteps = max(intSteps), interval = substr(hm(interval[intSteps ==maxSteps]/100),1,6))
maxActivity 
           
```
which is an average of  `r maxActivity$maxSteps` steps at `r maxActivity$interval`   time interval.


####_________________________________________________________________________
    
####    

    
####4.- Imputing missing values

a.- Calculating the number of missing values...

```{r}
dimTot <-dim(activity)

act_nas <-activity %>%
     filter(is.na(steps))
dimNas<-dim(act_nas)

```
... and, there are  `r dimNas[1]` NA's on a total of `r dimTot[1]` records  
  
Now creating a dataset with the NA's filled with the average value for that interval. We'll take the average from trim_activity that is groupped by interval and sumarized by it. 
So trim_activity holds the average value for each 5 minute period of the day. 
```{r}
corrected <-  activity %>%
            mutate(steps = ifelse(is.na(steps), trim_activity[trim_activity$interval == interval,]$intSteps, steps))%>% # ifelse() inserts on 'steps' the average value if the original value is NA
            filter(steps != "NA") ## filtering out rows with NA's on steps ( that are redundant since we inseted the averages)
newSteps <- corrected %>% ## now 'corrected' holds valid values, so  we do the same as in the 1st histogram case 
     group_by(date) %>% ## grouping by date
     summarise( daySteps = sum(steps)) ## and summing the steps by groupped date.
## histogram 
hist(newSteps$daySteps,col="darkblue", breaks =seq(0,25000, by = 1000), main = "Total steps per day",, xlab= ("Total daily Steps"), ylab= ("Nr. of occurences"))

newmean <- summarize(newSteps, mean(daySteps))
nwemedian <- summarize(newSteps, median(daySteps))

diff_median <- 100*abs(median-nwemedian)/median
diff_mean <- 100*abs(mean-newmean)/mean
diff_median
diff_mean
```
We can see that inserting the average value of the interval into NA values practically doesn't modify the chacracteristics of the disteibution, since
the mean is unaffected, obvious, since we have introduced mean values, and the median is affected by only `r diff_median`%, a negligible difference. 
The only concern is about stationality ( weekends vs non weekends)


####_________________________________________________________________________
    
####    

    
####5.- Are there differences in activity patterns between weekdays and weekends?

```{r fig.height =7}
weekactivity <- corrected %>%
                mutate( day = ifelse(wday(ymd(date))==7 | wday(ymd(date))==1, "weekend", "weekday" )) %>% ## adding a day category, then grouping and summarizing
                group_by(day,  interval) %>%
                summarise( intSteps = mean(steps))
             
par(mfrow = c(2, 1)) ## define a dual pannel, 2 rows, by 1.
par(mar = c(4,5,2,1))  ## defining margins.
## graphs code.
with(subset(weekactivity, day=="weekend"), { ## subsetting weekend days
    plot( interval/100, intSteps,type="l",  xlab=" Time of the Day(24 hour)", ylab=" Steps", ylim = range(0:230), col ="darkmagenta", xlim = range(0:24), xaxt="n") 
     })
    axis(1, at=c(2,4,6,8,10,12,14,16,18,20,22,24)) 
    title(main = "Average step activity by time inteval- Weekends") 
    grid()
   with(subset(weekactivity, day=="weekday"), { ## subsetting weekdays
    plot( interval/100, intSteps,type="l",  xlab=" Time of the Day(24 hour)", ylab=" Steps", ylim = range(0:230), col ="blue", xlim = range(0:24), xaxt="n") 
     })
    axis(1, at=c(2,4,6,8,10,12,14,16,18,20,22,24)) 
    title(main = "Average step activity by time inteval- Weekdays") 
    grid()



```
  
We can see a different pattern during weekdays compared to weekends, as expected. Where on weekdays we see an increase in the peak activity from 8 to 9:30, on weekend days we see an increase of the activity roughly evenly distributed from 8 to 20, so it seems the subject has a sedentary job but seems to enjoy sporty weekend days.









